Physcraper: a python package for continual update of evolutionary estimates using the Open Tree of Life

### 2020-08-18 docs/submission-mee-2020-08-18.tex
discussion

\textbf{Word count}: 3745

@@

The most recent \emph{Ilex} tree from Yao \emph{et al.} (\protect\hyperlink{ref-yao2020phylogeny}{2020}), has been made available in the \href{https://tree.opentreeoflife.org/curator/study/view/ot_1984}{OpenTree Phylesystem} and in the \href{https://datadryad.org/stash/dataset/doi:10.5061/dryad.k0p2ngf4x}{DRYAD repository}. The ``Yao2020'' tree, is the best sampled phylogenetic tree yet available for the hollies, with 175 tips.
This makes it a great case to highlight the functionalities of Physcraper and assess its limitations.
A tutorial as well as illustrated examples of functions implemented on each step of the analysis are available in Physcraper's documentation website.
Figure \ref{fig:results} shows results from a Physcraper analysis applied to an alignment of internal transcribed spacer DNA regions (ITS) from Gottlieb \emph{et al.} (\protect\hyperlink{ref-gottlieb2005molecular}{2005}). Physcraper ran on a local BLAST database, on a laptop Linux computer for 19hrs 45min to finish BLAST and RAxML analyses. Bootstrap analyses took an additional 13hrs.
The updated Gottlieb2005 tree contains all 41 distinct taxa from the original study plus 231 new tips, contributing phylogenetic data to 84 additional taxa within \emph{Ilex}. The tree is 99\% resolved, however, a large portion of branches are negligibly small (133 branches \textless{} 0.00001 substitution rate units), appearing as unresolved nodes (Fig. \ref{fig:results}).
For comparison, the Yao2020 tree also contains all 41 distinct taxa from the original Gottlieb2005 study,
and contributes phylogenetic data to 134 additional taxa within \emph{Ilex}, from which
67 are also in updated Gottlieb2005.

\begin{figure}
\includegraphics[width=1\linewidth]{docs/figs/ilex-results} \caption{A phylogenetic tree is the main result of a Physcraper analysis.}\label{fig:results}
\end{figure}

@@

Figure \ref{fig:results} caption continued: The tree results from updating an alignment that was used to generate a tree of 48 tips of the genus Ilex, published in Gottlieb et al.~2005 (black tips). Physcraper added 231 new sequences (red tips) corresponding to 84 distinct taxa that were not sampled in the original alignment. Physcraper obtained sequences from the GenBank database via BLAST, filtered them following specified criteria, aligned them to the original alignment using MUSCLE and performed a phylogenetic reconstruction using RAxML, as well as a conflict analysis using the OpenTree tools.

>>>>

\textbf{Word count}: 3803

@@

The most recent \emph{Ilex} tree from Yao \emph{et al.} (\protect\hyperlink{ref-yao2020phylogeny}{2020}), has been made available in the \href{https://tree.opentreeoflife.org/curator/study/view/ot_1984}{OpenTree Phylesystem} and in the \href{https://datadryad.org/stash/dataset/doi:10.5061/dryad.k0p2ngf4x}{DRYAD repository}. The ``Yao2020'' tree, is the best sampled phylogenetic tree yet available for the hollies, with 175 tips.
This makes it a great case to highlight the functionalities of Physcraper and assess its limitations.
A tutorial as well as illustrated examples of functions implemented on each step of the analysis are available in Physcraper's documentation website.
Figure \ref{fig:results} shows results from a Physcraper analysis applied to an alignment of internal transcribed spacer DNA regions (ITS) from Gottlieb \emph{et al.} (\protect\hyperlink{ref-gottlieb2005molecular}{2005}). Physcraper ran on a local BLAST database, on a laptop Linux computer for 19hrs 45min to finish BLAST and RAxML analyses. Bootstrap analyses took an additional 13hrs.
The updated Gottlieb2005 tree contains all 41 distinct taxa from the original study plus 231 new tips, contributing phylogenetic data to 84 additional taxa within \emph{Ilex}. The tree is 99\% resolved, however, a large portion of branches are negligibly small (133 branches \textless{} 0.00001 substitution rate units), appearing as unresolved nodes (Fig. \ref{fig:results}).
For comparison, the Yao2020 tree also contains all 41 distinct taxa from the original Gottlieb2005 study,
and contributes phylogenetic data to 134 additional taxa within \emph{Ilex}, from which
67 are also in updated Gottlieb2005. Yao \emph{et al.} (\protect\hyperlink{ref-yao2020phylogeny}{2020}) also used ITS as a marker, but their data in GenBank is not public yet, so Physcraper was unable to incorporate 68 additional taxa into the analysis. The Yao2020 tree also misses 18 taxa that were added by Physcraper. This might be related to the method they used to download existing ITS \emph{Ilex} sequences from GenBank, but they do not clarify this in their publication.

\begin{figure}
\includegraphics[width=1\linewidth]{docs/figs/ilex-results} \caption{A) Phylogenetic tree obtained by updating the Gottlieb et al. 2005 tree in B) using Physcraper.}\label{fig:results}
\end{figure}

@@

Figure \ref{fig:results} caption continued: Tips in original alignment and new tips added with Physcraper are depicted in black and red, respectively. Physcraper obtained sequences from the GenBank database via local BLAST of all sequences in the original alignment that generated tree in B), filtered them following criteria specified in section ``DNA sequence search and filtering'', aligned them to the original alignment using MUSCLE and performed a phylogenetic reconstruction using RAxML with 100 bootstraps. B-D show results of the conflict analysis performed with OpenTree tools.
